---
title: "Graphics with ggplot2"
author: "Stephanie"
date: "10/22/2018"
output: html_document
---

# learning ggplot2
```{r, echo=FALSE}

#install.packages('tidyverse')
library(tidyverse)

#National Parks in California
ca <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/ca.csv") 

#Acadia National Park
acadia <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/acadia.csv")

#Southeast US National Parks
se <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/se.csv")

#2016 Visitation for all Pacific West National Parks
visit_16 <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/visit_16.csv")

#All Nationally designated sites in Massachusetts
mass <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/mass.csv")

```

Look at dataset

```{r}
head(ca)
View(ca) #produces a little spreadsheet within R
```

# Single plots

1. tell ggplot what data you're using
2. create a geom (what type of plot) with aes (where should the actual points be)

```{r}
ggplot(data=ca) +
  geom_point(aes(x=year, y=visitors))
```

```{r}
ggplot(data=ca) +
  geom_point(aes(x=year, y=visitors, color=park_name))
```

```{r}
ggplot(data=ca) +
  geom_point(aes(x=year, y=visitors, color=park_name)) +
  labs(x = "Year",
       y = "Visitation",
       title = "California National Park Visitation") +
  theme_bw() + #changed background theme to a different one
  theme(legend.title = element_blank()) #removed title from legend
```

```{r}
ggplot(data=se) +
  geom_point(aes(x=year, y = visitors, color=park_name))
```

```{r}
ggplot(data=se) +
  geom_point(aes(x=year, y=visitors, color=state)) +
  xlab("Year") + 
  ylab("Visitation") +
  ggtitle("Southeast National Park Visitation") +
  theme_classic() +
  theme(legend.title = element_blank(),
        axis.text.x = element_text(angle=45, size=12, hjust = 1))
```

# Faceting plots

This makes subplots

```{r}
ggplot(data=se) +
  geom_point(aes(x=year, y=visitors)) +
  facet_wrap(~ state)
```

Free scales from each other. Also note that the color applies to the entire aes

```{r}
ggplot(data=se) +
  geom_point(aes(x=year, y=visitors, color=park_name)) +
  facet_wrap(~ state, scales="free")
```

Different geoms will give you different types of graphs. For things that are not dependent on the data, change options within the geom but outside the aes.

```{r}
ggplot(data = se) +
  geom_jitter(aes(x = park_name, y = visitors), alpha = 0.4, width = 0.1, color="blue") +
  coord_flip()
```

You can layer multiple geoms.

```{r}
ggplot(data = acadia) +
  geom_point(aes(x = year, y = visitors)) +
  geom_line(aes(x = year, y = visitors)) +
  geom_smooth(aes(x = year, y = visitors))
```

OR a more succient code

```{r}
ggplot(data = acadia, aes(x = year, y = visitors)) +
  geom_point() +
  geom_line() +
  geom_smooth()
```

You can also define data within the geom if you're overlaying from different datasets.

# Bar charts

Bar wants only an x and then it will count how many of each x. In order to use y as the height of each type in x, you have to do the stat = "identity"
```{r}
ggplot(data = visit_16, aes(x = state, y = visitors)) +
  geom_bar(stat = "identity")
```

```{r}
ggplot(data = visit_16, aes(x = state, y = visitors, fill = park_name)) +
  geom_bar(stat = "identity")
```

# Saving ggplots 
```{r}
myplot <- ggplot(data = visit_16, aes(x = state, y = visitors, fill = park_name)) +
  geom_bar(stat = "identity", position = "dodge")

myplot

ggsave(filename = "myplot.png", myplot)
```

