---
title: "Data wrangling"
author: "Stephanie"
date: "10/22/2018"
output: html_document
---

##Data wrangling

Looking at our data

```{r}
library(tidyverse)  # install.packages("tidyverse")

# read gapminder data csv. Note the readr:: prefix identifies which package it's in
gapminder <- readr::read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv')

# inspect the data
gapminder
head(gapminder)
tail(gapminder, 10)

# summary statistics
str(gapminder)
summary(gapminder)
skimr::skim(gapminder)  # install.packages("skimr")

# more...
names(gapminder)  #column names
dim(gapminder)
ncol(gapminder)  # or nrow
```

#dplyr

Filter

```{r}
# Filter by rows
filter(gapminder, lifeExp < 29)
filter(gapminder, country == "Mexico")
filter(gapminder, country %in% c("Mexico", "Peru"))

filter(gapminder, country == "Mexico", year == 2002)

# mean life expectancy in Sweden
swedes = filter(gapminder, country == "Sweden")
mean(swedes$lifeExp)  # $ looks at a column in a dataframe
```

Select

```{r}
# Select by columns
select(gapminder, year, lifeExp)
select(gapminder, -continent, -lifeExp)  # exclude columns
```

Piping!

```{r}
gap_cambodia <- filter(gapminder, country == "Cambodia") %>% select(-continent, -lifeExp)
gap_cambodia_alt <- gapminder %>% filter(country == "Cambodia") %>% select(-continent, -lifeExp)

# convention is
gap_cambodia_alt2 <- gapminder %>%
  filter(country == "Cambodia") %>%
  select(-continent, -lifeExp)
```

Mutate
Note that we use = inside functions; distinct from the <- to assign variables

```{r}
# mutate adds new variables

gapminder %>%
  mutate(index = 1:nrow(gapminder))  # created a new column called index that has numbers from 1 to the length of gapminder

gapminder %>%
  mutate(gdp = pop * gdpPercap)

gapminder %>% 
  filter(country %in% c("Egypt", "Vietnam")) %>%
  mutate(gdp = pop * gdpPercap,
         max_gdp = max(gdp))  # can mutate two columns within mutate
```

Groupby - make sure you ungroup at the end!

```{r}
# with mutate()
gap_grouped <- gapminder %>% 
  group_by(country) %>% 
  mutate(gdp = pop * gdpPercap,
         max_gdp = max(gdp)) %>% 
  ungroup()

# with summarize()
gap_summarized <- gapminder %>% 
  group_by(country) %>% 
  mutate(gdp = pop * gdpPercap) %>% 
  summarize(max_gdp = max(gdp)) %>%
  ungroup()
```

Arrange

```{r}
# arranges columns
gap_summarized <- gapminder %>% 
  group_by(country) %>% 
  mutate(gdp = pop * gdpPercap) %>% 
  summarize(max_gdp = max(gdp)) %>%
  ungroup() %>% 
  arrange(desc(max_gdp))  # descending order
```

Joining datasets
(helpful graphics in book)

```{r}
# read in the data. (same URL as yesterday, with co2.csv instead of gapminder.csv)
co2 <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/co2.csv")

# explore
co2 %>% head()
co2 %>% dim()  # 12 observations

# create a new gapminder variable with only 2007 data
gap_2007 <- gapminder %>% 
  filter(year == 2007)

gap_2007 %>% dim()  # 142 observations

# left join
lj <- left_join(gap_2007, co2, by = "country")  # 142 observations
# or
lj <- gap_2007 %>% 
  left_join(co2, by = "country")

# right join
rj <- right_join(gap_2007, co2)  # if you don't specify what to join by, it will guess
```

#tidyr

```{r}
# read in gapminder data in wide format
gap_wide <- readr::read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder_wide.csv')
```

Gather

```{r}
gap_long <- gap_wide %>% 
  gather(key = obstype_year,
         value = obs_values,
         -continent, -country)
```

Separate

```{r}
gap_long <- gap_wide %>% 
  gather(key = obstype_year,
         value = obs_values,
         -continent, -country) %>%
  separate(col = obstype_year,
           into = c("obs_type","year"),
           sep = "_",
           convert=TRUE)
  # year was read as a character because it was split from a character; that's why we use convert
```

Plot long data

```{r}
# plot life expectancy in Canada
canada_df <- gap_long %>% 
  filter(country == "Canada", obs_type == "lifeExp")

ggplot(data = canada_df, aes(x = year, y = obs_values)) +
  geom_line()


# plot life expectancy of all countries in the Americas
life_df <-gap_long %>% 
  filter(obs_type == "lifeExp",
         continent == "Americas")

ggplot(data = life_df, aes(x = year, y = obs_values, color = country)) +
  geom_line()


# plot mean life expectancy for each continent over time from 1982 to 2007
mean_life <- gap_long %>%
  filter(year >= 1982,
         year <= 2007,
         obs_type == "lifeExp") %>%
  group_by(year, continent) %>% 
  summarize(mean_lifeExp = mean(obs_values)) %>% 
  ungroup()

ggplot(data = mean_life, aes(x = year, y = mean_lifeExp, color = continent)) +
  geom_line() +
  labs(x = "Year",
       y = "Age (years)",
       title = "Life Expectancy by Continent, 1982-2007") +
  theme_bw() +
  theme(legend.title = element_blank())
```

Spread - turn long data into wide data

```{r}
gap_normal <- gap_long %>% 
  spread(key = obs_type, value = obs_values)

# recreate the gap_wide dataframe
gap_wide2 <- gap_long %>% 
  unite(col = obstype_year, c(obs_type, year), sep = "_") %>% 
  spread(key = obstype_year, value = obs_values)
```

